---
title: "Momspi_locom_testing"
author: "Jacob T. Nearing Ph.D."
date: "2023-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## R Setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
library(LOCOM)
library(ggforce)
```

```{r ggplot}
theme_set(theme_cowplot())

colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

## Setup MOMSPI data

Source setup file in _code and use the momspi_raw data that is not filtered (so we can get diversity estimates)

We can filter the data afterward.
```{r}
source(here::here("notebook/_code/momspi-setup.R"))

momspi_raw <- adjust_dirchlet(momspi_raw)

```

### Calibrate

```{r}
momspi_obs <- momspi

#calibrate to "raw" sequence counts as this is the required input for LOCOM
momspi_cal <- momspi %>% calibrate(bias_all_vec, norm = "none") %>% transform_sample_counts(function (x) round(x, 0))
```

### Compute diversity indices and divide data

```{r}
div <- list(
  'obs' = momspi_obs,
  'cal' = momspi_cal) %>%
  map(estimate_richness, measures = c('Shannon', 'InvSimpson')) %>%
  map_dfr(as_tibble, rownames = '.sample', .id = 'type') %>%
  transmute(
    .sample, type,
    div_q1 = exp(Shannon),
    div_q2 = InvSimpson,
  ) %>%
  pivot_longer(starts_with('div_'), names_to = 'index', values_to = 'diversity')

div_wide <- div %>%
  pivot_wider(values_from = diversity, names_from = c('index', 'type')) 
```

```{r}
div_wide %>%
  ggplot(aes(x = div_q1_obs)) +
  scale_x_log10() +
  annotate('rect', xmin = 0, xmax = 1.5, ymin = -Inf, ymax = Inf, 
    fill = 'blue', alpha = 0.3) +
  annotate('rect', xmin = 1.5, xmax = 3, ymin = -Inf, ymax = Inf, 
    fill = 'grey', alpha = 0.3) +
  annotate('rect', xmin = 3, xmax = Inf, ymin = -Inf, ymax = Inf, 
    fill = 'green', alpha = 0.3) +
  geom_histogram() +
  geom_vline(xintercept = c(1.5, 3), color = 'black')
```

```{r}
div_wide <- div_wide %>%
  mutate(
    div_group = case_when(
      div_q1_obs < 1.5 ~ 'Low',
      div_q1_obs > 3 ~ 'High',
      TRUE ~ 'Medium'
    ) %>% factor(levels = c('Low', 'Medium', 'High'))
  )
```

```{r}
momspi_div_obs <- momspi_obs %>%
  left_join_sample_data(div_wide, by = '.sample') %>%
  filter_sample_data(div_group %in% c('Low', 'High'))


momspi_div_cal <- momspi_cal %>%
  left_join_sample_data(div_wide, by = '.sample') %>%
  filter_sample_data(div_group %in% c('Low', 'High'))


taxa_to_test <- momspi_cal %>%
    transform_sample_counts(close_elts) %>%
    filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
    taxa_names %>%
    print

taxa_to_test <- gsub("-", ".", taxa_to_test)
```



# Running LOCOM

## Testing package tutorial
```{r}
library(LOCOM)
data("throat.otu.table")
data("throat.meta")
data("throat.otu.taxonomy")

# filtering out three samples with antibiotic use
filter.out.sam <- which(throat.meta$AntibioticUsePast3Months_TimeFromAntibioticUsage != "None")
throat.otu.table.filter <- throat.otu.table[-filter.out.sam,]
throat.meta.filter <- throat.meta[-filter.out.sam,]
Y <- ifelse(throat.meta.filter$SmokingStatus == "NonSmoker", 0, 1)
C <- ifelse(throat.meta.filter$Sex == "Male", 0, 1)

# filtering out rare OTUs
prop.presence <- colMeans(throat.otu.table.filter > 0)
filter.out.otu <- which(prop.presence < 0.2)
if (length(filter.out.otu) > 0) {
    throat.otu.table.filter <- throat.otu.table.filter[, -filter.out.otu]
    prop.presence <- prop.presence[-filter.out.otu]
}

# running locom
res <- locom(otu.table = throat.otu.table.filter, Y = Y, C = C, seed = 1, n.cores=4, n.perm.max = 1000)
res$p.global
res$detected.otu
```

## Observed data
```{r}
trait_obs <- momspi_div_obs@sam_data$div_group
trait_obs <- ifelse(trait_obs=="Low", 1, 0)
otu_tab_obs <- data.frame(momspi_div_obs@otu_table)


table(trait_obs)
dim(otu_tab_obs)

test <- sample(c(0:1), length(trait_obs), replace=TRUE)

#it seems locom doesn't run without a covariate?
obs_res <- locom(otu.table = otu_tab_obs, Y=trait_obs, C=test, seed=1, n.cores = 2, n.perm.max = 1000)

locom_effect_size_obs <- obs_res$effect.size[match(taxa_to_test, colnames(obs_res$effect.size))]
names(locom_effect_size_obs) <- taxa_to_test
```

## Calibrated data

```{r}
trait_cal <- momspi_div_cal@sam_data$div_group
trait_cal <- ifelse(trait_cal=="Low", 1, 0)
otu_tab_cal <- data.frame(momspi_div_cal@otu_table)


table(trait_cal)
dim(otu_tab_cal)

#test <- sample(c(0:1), length(trait), replace=TRUE)

#it seems locom doesn't run without a covariate?
cal_res <- locom(otu.table = otu_tab_cal, Y=trait_cal, C=test, seed=1, n.cores = 2, n.perm.max = 1000)

locom_effect_size_cal <- cal_res$effect.size[match(taxa_to_test, colnames(cal_res$effect.size))]
locom_effect_size_cal
names(locom_effect_size_cal) <- taxa_to_test
```

It seems that calibration causes some samples to be removed? Need to look into this deeper...

```{r}
plotting_frame <- data.frame(sample=taxa_to_test,
                             estimate=c(locom_effect_size_cal,locom_effect_size_obs),
                             type=c(rep("cal", 30), rep("obs", 30)))


#take negative of estimate to flip sign to match original analysis...
plotting_frame %>% mutate(
      .sample_fct = fct_reorder(sample, -estimate, .fun = max) %>%
        fct_relabel(function(x) str_replace_all(x, '_', ' '))
    ) %>% 
  ggplot(aes(y=.sample_fct, x=-estimate, color=type)) +
  geom_vline(xintercept = 0, color="grey") +
  geom_point(size=1.5) +
  theme_minimal_hgrid() +
  xlab("Locom effect size") +
  ylab("Species-level OTU")
```

Locom doesn't return effect size deviations so we cannot calculate standard error out of the box.

