
```{r}
set.seed(42)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  echo = FALSE
  #> out.width = "70%",
  #> fig.align = 'center',
  #> fig.width = 6,
  #> fig.asp = 0.618,
  #> fig.show = "hold"
)

#> options(dplyr.print_min = 6, dplyr.print_max = 6)
#> options(crayon.enabled = FALSE)

library(tidyverse)
library(here)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot(12))

close_elts <- function(x) x / sum(x)
```


```{r}
bias <- tibble(taxon = c("Lactobacillus", "Gardnerella", "Atopobium", "Spike-in"),
  efficiency = c(4.68, 0.16, 0.285, 1)) %>%
  mutate(across(efficiency, ~. / min(.)))
# efficiency associated with the targeted measurement of Lactobacillus
targeted_efficiency <- 2
a0 <- tribble(
  ~taxon, ~timepoint, ~abundance,
  #> "Lactobacillus", "T1", 5,
  #> "Lactobacillus", "T2", 1,
  #> "Gardnerella", "T1", 1,
  #> "Gardnerella", "T2", 10,
  #> "Atopobium", "T1", 3,
  #> "Atopobium", "T2", 1.5,
  "Lactobacillus", "T1", 5,
  "Lactobacillus", "T2", 0.5,
  "Gardnerella", "T1", 2,
  "Gardnerella", "T2", 8,
  "Atopobium", "T1", 3,
  "Atopobium", "T2", 1.5,
) %>%
  # Shrink abundance to have similar scale as proportions
  mutate(across(abundance, ~ . / max(.))) %>%
  left_join(bias, by = "taxon") %>%
  with_groups(timepoint, mutate,
    total_abundance = sum(abundance),
    proportion = close_elts(abundance),
    biased_abundance = efficiency * abundance,
    biased_proportion = close_elts(biased_abundance),
    biased_count = biased_proportion * 1e3,
    abundance_estimate_bulk = biased_proportion * total_abundance,
    #> abundance_estimate_spikein = biased_count / biased_count[taxon == "Spike-in"],
    abundance_estimate_targeted = biased_count * targeted_efficiency *
      abundance[taxon == "Lactobacillus"] / biased_count[taxon == "Lactobacillus"],
  )
a1 <- a0 %>%
  select(-total_abundance) %>%
  pivot_longer(-c(taxon, efficiency, timepoint), names_to = "type")
#> what are the fold changes?
a1 %>% 
  pivot_wider(names_from = timepoint) %>%
  mutate(fc = T2 / T1) %>%
  #> filter(str_detect(type, "proportion")) %>%
  arrange(type, desc(taxon)) %>%
  print(n = Inf)

a1 %>% 
  with_groups(c(type, timepoint), summarize, across(value, sum)) %>%
  pivot_wider(names_from = timepoint) %>%
  mutate(fc = T2 / T1)
```

For now, focus on plotting the proportions. 
Show the error in the top panel: Actual -> Observed, for each time point. 
Show the fold changes in the bottom panel: T1 -> T1, for both proportion types.

```{r}
ptb <- a1 %>%
  filter(
    type %in% c("proportion", "biased_proportion"),
  ) %>%
  mutate(
    across(type, fct_relevel, "proportion", "biased_proportion"),
    across(type, fct_recode,
      "Actual" = "proportion",
      "Observed" = "biased_proportion"),
  )

shared_layers <- list(
  geom_path(aes(group = taxon), 
    arrow = grid::arrow(length = unit(0.15, "inches"))),
  geom_point(size = 2),
  scale_color_brewer(type = "qual", palette = 1, guide = FALSE),
    # guide = guide_legend(reverse = TRUE)),
  labs(y = "Proportion", color = "Taxon"),
  scale_y_log10(),
  coord_cartesian(clip = "off"),
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  expand_limits(y = 1e-2),
  theme(plot.margin = unit(c(0, 1.3, 0, 0), units = "in"))
)

# How much to nudge the taxon labels and proportions
nudge.taxon <- 0.48
nudge.prop <- 0.24
# In future iterations, consider labelling the taxa in both facets

p1 <- ptb %>%
  ggplot(aes(type, value, color = taxon)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  shared_layers +
  geom_text(data = ~filter(., type == "Actual"), 
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., type == "Observed"), 
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Type",
    title = "Measurement error at each time point"
  )
p2 <- ptb %>%
  ggplot(aes(timepoint, value, color = taxon)) +
  facet_wrap(~type, nrow = 1, scales = "fixed") +
  shared_layers +
  geom_text(data = ~filter(., timepoint == "T1"), 
    aes(label = round(value, 2)), nudge_x = -nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2"), 
    aes(label = round(value, 2)), nudge_x = nudge.prop) +
  geom_text(data = ~filter(., timepoint == "T2", type == "Observed"), 
    aes(label = taxon), nudge_x = nudge.taxon, hjust = 0) +
  labs(
    x = "Time point",
    title = "Actual and observed fold changes"
  )
p1 / p2
```

In addition, I want a plot showing the efficiencies and mean efficiencies, so that they can be related to the error.

First, compute the mean efficiency, then join with the taxon efficiencies in a table for plotting.
```{r}
sme <- a0 %>%
  with_groups(timepoint, summarize, mean_efficiency = sum(proportion * efficiency))
sme0 <- sme %>%
  mutate(label = str_glue("mean ({timepoint})")) %>%
  select(label, efficiency = mean_efficiency)
lvls = c("Atopobium", "Gardnerella", "Lactobacillus", "mean (T1)", "mean (T2)")
etb <- bind_rows(
  bias %>% filter(taxon != "Spike-in") %>% mutate(label = taxon, type = "Taxon"),
  sme0 %>% mutate(type = "Mean")
) %>%
  mutate(across(label, factor, levels = lvls))
```

Plot the individual and mean efficiencies. 
Set the vertical span to be 100 to match the other plots.
```{r}
#> RColorBrewer::display.brewer.all(type = "qual")
#> clrs <- RColorBrewer::brewer.pal(n = 8, "Accent")[c(1:3, 7:8)]
clrs <- c(RColorBrewer::brewer.pal(n = 3, "Accent"), rep("#585858", 2))
p3 <- etb %>%
  ggplot(aes(x = 1, y = efficiency, color = label)) + 
  geom_path(data = ~filter(., type == "Mean"), group = 1, 
    color = "black", size = 0.3, x = 1,
    arrow = grid::arrow(length = unit(0.15, "inches"))) +
  geom_point(aes(size = type, shape = type)) +
  geom_text(aes(label = label), nudge_x = 0.3, size = 5, hjust = 0) +
  scale_shape_manual(values = c(95, 16), guide = FALSE) +
  scale_size_manual(values = c(10, 2), guide = FALSE) +
  #> scale_color_brewer(type = "qual", palette = 1, guide = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
    title = "Taxonomic bias of protocol"
  ) +
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb$efficiency, 
    #> labels = round(etb$efficiency, 1)
    labels = signif(etb$efficiency, 2)
    ) +
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  #> expand_limits(y = 1e-2) +
  xlim(c(0.5, 3)) +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none"
  )
p3
```

Create a version split by time point,
```{r}
sme1 <- sme %>%
  mutate(taxon = "Mean", type = "Mean") %>%
  select(taxon, efficiency = mean_efficiency, timepoint, type)
bias1 <- bias %>% 
  filter(taxon != "Spike-in") %>% 
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(type = "Taxon")
lvls = c("Atopobium", "Gardnerella", "Lactobacillus", "Mean")
etb1 <- bind_rows(bias1, sme1) %>%
  mutate(across(taxon, factor, levels = lvls))
```
```{r}
p4 <- etb1 %>%
  ggplot(aes(x = 1, y = efficiency, color = taxon)) + 
  geom_point(aes(size = type, shape = type)) +
  geom_text(aes(label = taxon), nudge_x = 0.3, size = 5, hjust = 0) +
  scale_shape_manual(values = c(95, 16), guide = FALSE) +
  scale_size_manual(values = c(10, 2), guide = FALSE) +
  #> scale_color_brewer(type = "qual", palette = 1, guide = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
    title = "Taxonomic bias of protocol"
  ) +
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb1$efficiency, 
    #> labels = round(etb$efficiency, 1)
    labels = signif(etb1$efficiency, 2)
    ) +
  # scale_y_log10(breaks = c(1e-2, 3e-2, 1e-1, 3e-1, 1)) +
  #> expand_limits(y = 1e-2) +
  xlim(c(0.5, 2)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none"
  )
p4
```

Try a different way.
```{r}
lvls = c("Gardnerella", "Atopobium", "Lactobacillus")
bias2 <- bias %>% 
  filter(taxon != "Spike-in") %>% 
  expand(nesting(taxon, efficiency), timepoint = c("T1", "T2")) %>%
  mutate(
    type = "Taxon",
    across(taxon, factor, levels = lvls),
    x = as.integer(taxon)
  ) %>%
  left_join(sme, by = "timepoint")

```
```{r}
p4 <- bias2 %>%
  #> mutate(
  #>   across(timepoint, factor, 
  #>     levels = c("T1", "T2"), 
  #>     labels = str_c("Time point T", 1:2))
  #> ) %>%
  ggplot(aes(x = x, y = efficiency, color = taxon)) + 
  geom_point(size = 2) +
  geom_text(data = ~filter(., timepoint == "T2"),
    aes(label = taxon), x = 3.3, hjust = 0) +
  geom_segment(aes(xend = x, yend  = efficiency, y = mean_efficiency),
    arrow = grid::arrow(length = unit(0.15, "inches"))) +
  geom_segment(data = sme, 
    aes(x = 1, xend = 3, y = mean_efficiency, yend = mean_efficiency), 
    color = "#585858",
    inherit.aes = FALSE) +
  #> geom_text(data = sme0 %>% filter(timepoint == "T2"), 
  geom_text(data = sme0 %>% mutate(timepoint = "T2"),
    aes(label = label, y = efficiency), x = 3.3, hjust = 0,
    color = "#585858",
    inherit.aes = FALSE) +
  scale_color_manual(values = clrs) +
  labs(y = "Relative efficiency", x = NULL, color = "Taxon", 
    title = "Taxonomic bias of protocol"
  ) +
  scale_y_log10(limits = c(1, 100) / 2,
    breaks = etb1$efficiency, 
    labels = signif(etb1$efficiency, 2)
    ) +
  xlim(c(0.0, 4)) +
  facet_wrap(~timepoint, nrow = 1, scales = "fixed", 
    labeller = as_labeller(function(x) str_c("Time point ", x))
  ) +
  coord_cartesian(clip = "off") +
  theme(
    axis.line.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    plot.margin = unit(c(0, 1.0, 0, 0), units = "in")
  )
p4
```


### Multipanel


```{r}
p1 + p4 + p2 + plot_spacer() + 
  plot_layout(byrow = TRUE, ncol = 2, widths = c(1, 0.6)) +
  plot_annotation(tag_levels = "A")
```

```{r}
str_c("figures/example-vaginal-microbiome.", fmts) %>%
  here %>%
  walk(ggsave, units = "in", width = 8, height = 6, scale = 1.5)
```

This is a nice start.
The difference between a taxon's efficiency and the mean is the error for that sample.
It could make sense to show this with arrows. 


Next:

- Experiment with ways to make the connections even clearer between the efficiency differences and the error. 
- Try making the arrows vertical, so that they are directly connected to the fold changes in proportion and in efficiency.
- Can also split the third plot into T1 and T2.
- Might consider changing "Time point T1" to "Sample S1", to fit the primary language of the text.
- Can mention in the text that these could be samples from different women, or the same woman at different times.
- also use a shape scale for the taxa, to give a second way besides color to distinguih them


## Notes about the example

In this example, the proportion of Atopobium is observed to increase; it actually decreases.
Since the total abundance increases, the total abundance of Atopobium is also incorrectly observed to increase.

