---
title: "MOMS-PI: Key dynamics"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:brooks2015thet
  - ref:fettweis2019thev
date: 2021-11-01
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

This document is part of a series of analyses of the impact of taxonomic bias on vaginal community dynamics in MOMSPI dataset (@fettweis2019thev), using the bias estimated from the @brooks2015thet mock community data.
The purpose of this document is to demonstrate the key dynamics we want to highlight in the manuscript.

# Setup

## R setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

```{r ggplot}
theme_set(theme_cowplot())

colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

## Estimate bias from @brooks2015thet

We can estimate the efficiencies for the control species used in the cellular mocks of @brooks2015thet by running the example code in the docs for `metacal::estimate_bias()`. 
This code uses the observed and actual species-level abundance tables which were produced by @mclaren2019cons from the SI files of @brooks2015thet and are included in the metacal package.

```{r}
dr <- system.file("extdata", package = "metacal")
list.files(dr)
actual <- file.path(dr, "brooks2015-actual.csv") |>
  read.csv(row.names = "Sample") |>
  as("matrix")
observed <- file.path(dr, "brooks2015-observed.csv") |>
  read.csv(row.names = "Sample") |>
  subset(select = - Other) |>
  as("matrix")

# Estimate bias with bootstrapping for error estimation
mc_fit <- estimate_bias(observed, actual, margin = 1, boot = TRUE)
summary(mc_fit)

rm(actual, observed, dr)
```

```{r}
control_species <- mc_fit %>% coef %>% names
control_genera <- control_species %>% str_extract('^[^_]+')
```

## Load the MOMSPI data

TODO: replace this code chunk with one that pulls data from the github repo into `_data/`

Load the MOMSPI Stirrups profiles into phyloseq

```{r}
path_momspi <- '~/research/momspi'

otu <- path(path_momspi, "output", "stirrups-profiles", "abundance-matrix.csv.bz2") %>%
  read_csv(
    col_types = cols(.default = col_double(), sample_name = col_character())
  ) %>%
  otu_table(taxa_are_rows = FALSE)
sam <- path(path_momspi, "output", "stirrups-profiles", "sample-data.csv.bz2") %>%
  read_csv(col_types = "ccccccic") %>%
  mutate(across(host_visit_number, factor, ordered = TRUE)) %>%
  sample_data
tax <- path(path_momspi, "output", "stirrups-profiles", "taxonomy.csv.bz2") %>%
  read_csv(col_types = cols(.default = col_character())) %>%
  tax_table %>%
  mutate_tax_table(
    species = case_when(!is.na(genus) ~ .otu)
  )
momspi_raw <- phyloseq(otu, sam, tax) %>%
  mutate_tax_table(across(.otu, str_replace, 
      "(?<=Lactobacillus_crispatus)_cluster", "")) %>%
  mutate_sample_data(., sample_sum = sample_sums(.))
taxa_names(momspi_raw) %>% str_subset("crispatus")
stopifnot(all(control_species %in% taxa_names(momspi_raw)))
```

L crisp type 1 is just a few reads so we can ignore it (check)

```{r}
p1 <- momspi_raw %>% sample_sums %>% qplot + scale_x_log10() +
  labs(x = 'sample sum')
p2 <- momspi_raw %>% taxa_sums %>% qplot + scale_x_log10() +
  labs(x = 'otu sum')
p1 + p2
```

Let's filter samples with fewer than 1000 reads (also requirement used by the authors)

```{r}
momspi <- momspi_raw %>% 
  filter_sample_data(sample_sum >= 1e3)
```

Note, we might want to be conservative in terms of composition shifts by making the min prop of species 1e-3.
However this might add up to a large effect if we apply it to all species.
Could instead filter to 1e4 reads and threshold at 1e-4, or 3e-4

Let's see if we can filter the super-low-count species without affecting any remaining samples too much:

```{r}
x <- (momspi %>% filter_taxa2(~sum(.) >= 2e2) %>% sample_sums) / (momspi %>% sample_sums)
summary(x)
```

Dropping OTUs with less than 200 reads has a negligible impact on the samples, so let's do that.

```{r}
momspi <- momspi %>% filter_taxa2(~sum(.) >= 2e2)
```

Finally, let's impute zeros using the Dirichlet posterior mean, with the prior respecting the average proportions across samples.

```{r}
taxa_mean_prop <- momspi %>%
  otu_table %>%
  transform_sample_counts(close_elts) %>%
  orient_taxa(as = 'cols') %>%
  as('matrix') %>%
  colMeans
stopifnot(sum(taxa_mean_prop) == 1)
prior_vec <- taxa_mean_prop * ntaxa(momspi)
stopifnot(identical(length(prior_vec), ntaxa(momspi)))
# Note the need for the seq_along trick to get phyloseq to allow this
# adjustment
momspi <- momspi %>% transform_sample_counts(~ prior_vec[seq_along(.x)] + .x)
```

## Checks

how many subjects?

```{r}
cts <- momspi %>% sample_data %>% as_tibble %>% count(host_subject_id)
nrow(cts)
```

how does the number of visits vary by subject?

```{r}
cts %>% pull(n) %>% summary
cts %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 1)
```

# Calibrate the MOMSPI measurements

Owing to the large sample size of the mock community experiment, the standard errors in the estimated bias for the control species are quite small, and
our uncertainty in the true bias of the MOMSPI study is much larger than what these standard errors imply.
Therefore I will use the point estimate of the bias for the control species as if it were known exactly, but we should consider this to just be a hypothesis of what the bias truly is in the MOMSPI study.
This hypothetical nature applies even more strongly when we extrapolate to other species.

Since don't have a way to directly estimate the efficiencies of the other OTUs, we'll impute them as follows.

1. Set the efficiencies of the 7 control species to the directly estimated values
2. Compute efficiencies for the 6 control genera to the efficiency of the control species (if only one) or the geometric mean efficiency (if more than one; here this is just _Lactobacillus_)
3. Use these genera-level efficiencies for the efficiencies of OTUs that are not control species but are in the same genus as one
4. For other OTUs, use the geometric mean of the genera efficiencies; this is similar to using the mean of the 7 control species but gives _Lactobacillus_ equal weight to the other genera.

<!--  -->

```{r}
bias_species <- coef(mc_fit) %>% 
  enframe("species", "efficiency") %>%
  print
bias_genus <- bias_species %>%
  mutate(genus = str_extract(species, "^[^_]+"), .before = 1) %>%
  with_groups(genus, summarize, across(efficiency, gm_mean)) %>%
  print
# Match on genus or species, depending on which is available; then set others
# to average genus efficiency
bias_all <- tax_table(momspi) %>% as_tibble %>%
  left_join(bias_species, by = "species") %>%
  left_join(bias_genus, by = "genus") %>%
  mutate(
    efficiency = case_when(
      !is.na(efficiency.x) ~ efficiency.x,
      !is.na(efficiency.y) ~ efficiency.y,
      TRUE ~ gm_mean(bias_genus$efficiency)
    )
  ) %>%
  select(-efficiency.x, -efficiency.y) %>%
  # standardize to L. iners, the most efficiently measured
  mutate(
    across(efficiency, ~ . / max(.))
  ) %>%
  glimpse
bias_all_vec <- bias_all %>% select(.otu, efficiency) %>% deframe 
```

## Observed and calibrated proportions

Compute observed and calibrated proportions with all taxa.

```{r}
momspi_obs <- momspi %>% 
  transform_sample_counts(close_elts)
# note, calibrate() automatically subsets to just the focal taxa if we haven't
# already
momspi_cal <- momspi_obs %>% calibrate(bias_all_vec)
```

## Sample summary statistics

- Mean efficiencies
- Most abundant species
- Proportion of Lactobacillus
<!--  -->

To compute the mean efficiencies, we can use `metacal::perturb()` to multiply efficiencies by the actual proportions for each sample (using `norm = "none"` to avoid renormalizing the results), then summing up the values from each sample with `phyloseq::sample_sums()`.
It's also useful to get the mean efficiency among just the control species.

```{r}
tmp1 <- momspi_cal %>%
  perturb(bias_all_vec, norm = "none") %>% 
  sample_sums %>%
  enframe('.sample', 'mean_efficiency')
tmp2 <- momspi_cal %>%
  filter_tax_table(.otu %in% control_species) %>%
  transform_sample_counts(close_elts) %>%
  perturb(bias_species %>% deframe, norm = "none") %>% 
  sample_sums %>%
  enframe('.sample', 'mean_efficiency_controls')
mean_eff <- left_join(tmp1, tmp2, by = '.sample')
```

For each sample, let's get the most abundant species in the calibrated and uncalibrated profiles. 

```{r}
get_top_species <- function(ps) {
  x <- ps %>%
    otu_table %>%
    as_tibble(pivot = TRUE) %>%
    with_groups(.sample, slice_max, .abundance, n = 1) %>%
    # add count to check for ties
    add_count(.sample)
  if (anyDuplicated(x$.sample))
    stop('Ties present')
  x %>% select(.sample, .otu, .abundance)
}
top_species <- list(
  Uncalibrated = momspi_obs,
  Calibrated = momspi_cal
  ) %>%
  map_dfr(get_top_species, .id = 'type')
```

Which species tend to be dominant?

```{r}
top_species %>%
  filter(type == 'Uncalibrated') %>%
  count(.otu) %>%
  arrange(desc(n)) %>%
  print(n=20)
```

```{r}
top_species %>%
  filter(type == 'Calibrated') %>%
  count(.otu) %>%
  arrange(desc(n)) %>%
  print(n=20)
```

The most important species/OTU that we do not have direct bias information on at the species or genus level is Lachnospiraceae BVAB1.

The proportion of Lactobacillus in each sample - will simply do all control genera.

```{r}
genus_prop <- momspi_cal %>%
  subset_taxa(genus %in% control_genera) %>%
  tax_glom('genus') %>%
  as_tibble %>%
  select(.sample, genus, .abundance) %>%
  pivot_wider(names_from = genus, values_from = .abundance)
```

Let's join the most abundant species and key genera proportions to the mean efficiecies for further analysis.
For plotting, we'll create a version of the top taxa that only colors the taxa that are most often dominant.

```{r}
tmp <- top_species %>% 
  select(-.abundance) %>% 
  pivot_wider(names_from = type, values_from = .otu, 
    names_glue = 'top_{str_to_lower(type)}')
mean_eff1 <- mean_eff %>%
  left_join(sample_data(momspi_cal) %>% as_tibble, by = '.sample') %>%
  left_join(genus_prop, by = '.sample') %>%
  left_join(tmp, by = '.sample') %>%
  mutate(
    top_plot = fct_lump_n(top_calibrated, 7)
  )
rm(tmp)
```

# Visual analysis of mean efficiency distribution and role of key species

We can make a plot similar to that used for the Leopold et al analysis.

```{r}
# First let's look at the distribution of mean efficiencies
p1 <- mean_eff1 %>%
  ggplot(aes(mean_efficiency, fill = top_plot)) +
  geom_histogram() +
  # annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  expand_limits(x = range(bias_all_vec)) +
  scale_fill_brewer(type = "qual") +
  labs(
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency", 
    y = "Number of samples",
    fill = 'Most abundant species'
  )
# Next let's plot Lactobacillus proportion against mean efficiency, as another
  # look at the relationship between Lacto and mean efficiency.
p2 <- mean_eff1 %>%
  ggplot(aes(mean_efficiency, Lactobacillus, color = top_plot)) +
  geom_point() +
  scale_x_log10() +
  expand_limits(x = range(bias_all_vec)) +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 1e-2),
    breaks = c(0, 0.01, 0.03, 0.1, 0.3, 1)
    ) +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Lactobacillus proportion",
    x = "Mean efficiency",
    color = 'Most abundant species'
  )
```

```{r momspi-mean-efficiency-dist-and-lacto, fig.dim = c(6,5) *1.6}
# The yellow makes the L crispatus points a bit hard to see, so let's swap it's
# color with the rarer L. gasseri in the final figure.
colors_top <- RColorBrewer::brewer.pal(8, 'Accent')[c(1,2,3,5,4,6,7,8)] 
(p1 + ggtitle("Distribution of sample mean efficiencies")) / 
  (p2 + 
    ggtitle("Lactobacillus proportion vs. mean efficiency") + 
    theme(legend.position = 'none')
  ) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = 'collect') &
  scale_fill_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  ) &
  scale_color_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  )
# The x-axis is very close but not completely aligned; I think I need to
# manually set the x limits to get perfect alignment, but that leads to some
# samples being dropped in the histogram plot; some investigation is needed
# to go that route.
```

This plot shows that Lactobacillus dominance drives a bimodal distribution in the mean efficiencies, and that Gardnerella dominance drives the low end.
Note that the 'other' and 'BVAB1' dominated samples are lumped around the middle because of our assumption in imputing the efficiencies.

For the caption/methods: Note that the most abundant species is that inferred from the calibrated data; could just say that the calibrated data is treated as the actual composition.

The Lacto proportion doesn't seem that much more illuminating than the histogram colored by dominant species, so let's make a version with just that for the manuscript.

```{r momspi-mean-efficiency-dist, fig.dim = c(7,3.5)}
colors_top <- RColorBrewer::brewer.pal(8, 'Accent')[c(1,2,3,5,4,6,7,8)] 
p1 + 
#  ggtitle("Distribution of sample mean efficiencies") +
  scale_fill_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  ) +
  scale_color_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  )
```

# Variation in mean efficiency and individual species

```{r}
focal_taxa <- momspi %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
  taxa_names %>%
  print
```

```{r}
x <- momspi_cal %>%
  transform_sample_counts(close_elts) %>%
  filter_tax_table(.otu %in% focal_taxa) %>%
  as_tibble
```

```{r}
taxa_summary <- x %>%
  mutate(across(.abundance, log)) %>%
  with_groups(.otu, summarize,
    across(.abundance, 
      .fns = c(mean = mean, median = median, sd = sd, var = var),
      .names = '{.fn}'
    )
  )
taxa_summary %>% arrange(-mean) %>% print(n=Inf)
```

```{r}
me_summary <- mean_eff1 %>%
  mutate(across(mean_efficiency, log)) %>%
  summarize(
    across(mean_efficiency,
      .fns = c(sd = sd, var = var),
      .names = '{.fn}'
    )
  ) %>%
  print
```

The mean efficiency has a smaller dispersion than individual species, but it is of a similar magnitude, so there is still potential for it to have a substantial impact.

What about variation within women?

```{r}
taxa_summary_within <- x %>%
  mutate(across(.abundance, log)) %>%
  with_groups(c(.otu, host_subject_id), summarize,
    across(.abundance,
      .fns = c(mean = mean, median = median, sd = sd, var = var),
      .names = '{.fn}'
    ),
    n = n()
  ) %>%
  filter(n >= 3) %>%
  with_groups(.otu, summarize,
    sd_mean = sd(mean),
    mean_sd = mean(sd)
  )
taxa_summary_within %>% print(n=Inf)
```

```{r}
me_summary_within <- mean_eff1 %>%
  mutate(across(mean_efficiency, log)) %>%
  with_groups(host_subject_id, summarize,
    across(mean_efficiency,
      .fns = c(mean = mean, sd = sd, var = var),
      .names = '{.fn}'
    ),
    n = n()
  ) %>%
  filter(n >= 3) %>%
  summarize(
    sd_mean = sd(mean),
    mean_sd = mean(sd)
  ) %>%
  print
```

Note, a better way to do this would perhaps be to use anova / linear modeling, or whatever method from those would properly weight the number of samples from each woman.

what about the variation between the first samples from each woman?

```{r}
taxa_summary_first <- x %>%
  with_groups(host_subject_id, slice_min, host_visit_number, n = 1) %>%
  mutate(across(.abundance, log)) %>%
  with_groups(.otu, summarize,
    across(.abundance,
      .fns = c(mean = mean, median = median, sd = sd, var = var),
      .names = '{.fn}'
    )
  )
taxa_summary_first %>% arrange(-mean) %>% print(n=Inf)
# taxa_summary %>% arrange(-mean) %>% print(n=Inf)
```

These sd's are similar to the overall sds, which makes sense given most pairwise comparisons are between women.

```{r}
me_summary_first <- mean_eff1 %>%
  with_groups(host_subject_id, slice_min, host_visit_number, n = 1) %>%
  mutate(across(mean_efficiency, log)) %>%
  summarize(
    across(mean_efficiency,
      .fns = c(sd = sd, var = var),
      .names = '{.fn}'
    )
  ) %>%
  print
me_summary
```

This is somewhat larger, but again similar.

Tentative: 
We see that the (log-e) mean efficiency sd is around 0.9, while the sd of individual taxa ranges from 2-3.5 (between women and overall).
Within women, the mean efficiency sd and that of individual taxa are both smaller, such that the relative importance of each seems similar.

Visualize, for the first sample from each woman:

```{r}
bind_rows(
  x, 
  mean_eff1 %>% mutate(.otu = 'Mean efficiency', .abundance = mean_efficiency)
) %>%
  with_groups(c(.otu, host_subject_id), slice_min, host_visit_number, n = 1) %>%
  mutate(
    across(.abundance, log),
    across(.otu, fct_reorder, .abundance)
  ) %>%
  ggplot(aes(y = .otu, x = .abundance)) +
  ggdist::stat_slab()
```

# Fold changes in mean efficiency

How often do shifts in the mean efficiency occur within a single subject?
How often do shifts from Lacto to non-Lacto dominance occur within a single subject?
How often do shifts from Lacto to Gardnerella dominance occur within a single subject?

Let's start by computing the pairwise changes in the sample mean efficiency between consecutive samples within women, and between the first samples of different women (first samples only to limit the number of sample pairs).

Since we may be missing some host visits, let's create a new host visit number that has no gaps:

```{r}
mean_eff2 <- mean_eff1 %>%
  with_groups(host_subject_id, mutate, 
    host_visit_rank = rank(host_visit_number))
```

```{r}
tb <- crossing(
  sample.1 = mean_eff2$.sample, 
  sample.2 = mean_eff2$.sample
  ) %>%
  left_join(mean_eff2, by = c(sample.1 = '.sample')) %>%
  left_join(mean_eff2, by = c(sample.2 = '.sample'), suffix = c('.1', '.2')) %>%
  mutate(
    comparison_type = case_when(
      host_subject_id.1 == host_subject_id.2 & host_visit_rank.2 == host_visit_rank.1 + 1 ~ 
        'Between consecutive visits',
      host_subject_id.1 != host_subject_id.2 & host_visit_rank.2 == 1 & host_visit_rank.1 == 1 ~
        "Between subjects' first visits"
    )
  ) %>%
  filter(!is.na(comparison_type)) %>%
  mutate(
    switch_class = case_when(
      top_plot.1 == top_plot.2 ~ 'No change',
      str_detect(top_plot.1, 'Gardnerella') & str_detect(top_plot.2, 'Lactobacillus') ~ 'Gard. to Lacto.',
      str_detect(top_plot.1, 'Lactobacillus') & str_detect(top_plot.2, 'Gardnerella') ~ 'Lacto. to Gard.',
      str_detect(top_plot.1, 'Lactobacillus') & str_detect(top_plot.2, 'Lactobacillus') ~ 'Lacto. to Lacto.',
      TRUE ~ 'Other',
    )
  )

# tb_within <- tb %>%
#   filter(host_subject_id.1 == host_subject_id.2, host_visit_rank.2 == host_visit_rank.1 + 1)
# tb_between <- tb %>%
#   filter(host_subject_id.1 != host_subject_id.2, host_visit_rank.2 == 1, host_visit_rank.1 == 1)
```

Let's look at the distribution of FCs in mean efficiency, coloring by whether there is a shift in dominant species.

```{r momspi-mean-efficiency-fcs, fig.dim = c(6, 7.0)}
tb %>%
  ggplot(aes(mean_efficiency.2 / mean_efficiency.1, fill = switch_class)) +
  facet_wrap(~comparison_type, ncol = 1, scales = 'free_y') +
  geom_histogram() +
  scale_x_log10() +
  # expand_limits(x = range(bias_all_vec)) +
  labs(
    x = "Fold change in mean efficiency", 
    y = "Number of sample pairs",
    fill = 'Change in most\nabundant species'
  ) +
  scale_fill_brewer(type = 'qual', palette = 3)
```

Unsurprisingly, shifts in the mean efficiency tend to be larger between different subjects.
Note, the between-subject panel is perfectly symmetric, but the consecutive-visits panel is not. 

```{r}
tb %>%
  count(comparison_type, switch_class) %>%
  with_groups(comparison_type, mutate, prop = close_elts(n)) %>%
  knitr::kable(digits = 3)
```

# Example community trajectory

Pick a woman with multiple visits and a large change in mean efficiency to illustrate the potential impact of bias on species trajectories.
We expect large changes in efficiency to be drive by changes in dominance from Gard to Lactobacillus (or vice versa).

To pick a woman, let's first examine the distribution of geometric standard deviations of mean efficiency across subjects, and pick a subject at the upper end who has multiple visits.

```{r}
me <- mean_eff1 %>%
  mutate(across(mean_efficiency, log)) %>%
  with_groups(host_subject_id, summarize,
    across(mean_efficiency,
      .fns = c(mean = mean, sd = sd, var = var),
      .names = '{.fn}'
    ),
    n = n()
  ) %>%
  filter(n >= 3)
```

```{r}
me %>% slice_max(sd, n = 10)
```

Let's pick subject 'EP280575', who has the most visits within the top 10.

```{r}
sid <- 'EP280575'
ps_obs <- momspi %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~max(.) > 1e-3)
ps_cal <- momspi_cal %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  prune_taxa(taxa_names(ps_obs), .)
x <- list(Calibrated = ps_cal, Observed = ps_obs) %>%
  map_dfr(as_tibble, .id = "type") %>%
  mutate(across(.otu, fct_relabel, function(x) str_replace_all(x, '_', ' ')))
```

```{r, fig.dim = c(6, 15)}
taxa_to_plot <- x %>% 
  filter(type == 'Calibrated') %>%
  with_groups(.otu, summarize, 
    med = median(log10(.abundance)),
    max = max(log10(.abundance)),
  ) %>%
  filter(max > -2.5) %>%
  arrange(-max) %>%
  pull(.otu) %>%
  print
x %>%
  filter(.otu %in% taxa_to_plot) %>%
  mutate(.otu = factor(.otu, taxa_to_plot)) %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
  geom_line(aes(group = type)) +
  labs(y = 'Proportion', title = 'Species proportions') +
  scale_color_brewer(type = 'qual', palette = 6) +
  theme(legend.position = 'top')
```

For the final plot, I'll pick the most abundant 3 taxa (which are driving the mean efficiency) and two more taxa that show the different behaviors.

```{r trajectories-within-subject-high-var, fig.dim = c(6, 9)}
taxa_to_plot <- c(
  'Gardnerella_vaginalis', 'Lachnospiraceae_BVAB1', 'Lactobacillus_iners',
  'Ureaplasma_cluster23', 'Megasphaera_OTU70_type1') %>%
  str_replace_all('_', ' ')
p_sp <- x %>%
  filter(.otu %in% taxa_to_plot) %>%
  mutate(.otu = factor(.otu, taxa_to_plot)) %>%
  filter(as.integer(.otu) <= 6) %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
  geom_line(aes(group = type)) +
  labs(y = 'Proportion', title = 'Species proportions') +
  scale_color_brewer(type = 'qual', palette = 6) +
  theme(legend.position = 'top')
p_me <- mean_eff1 %>%
  filter(host_subject_id == sid) %>%
  ggplot(aes(host_visit_number, mean_efficiency)) +
  scale_y_log10() +
  labs(y = 'Efficiency', title = 'Mean efficiency') +
  geom_line(aes(group = 1))
p_me / p_sp &
  annotate('rect', xmin = 3, xmax = 4, ymin = Inf, ymax = 0, alpha = 0.1) &
  labs(x = 'Host visit') &
  plot_layout(heights = c(0.1, 1)) &
  plot_annotation(tag_levels = 'A')
```


For comparison, let's also look at a long trajectory where there is no major change in mean efficiency.

```{r}
me %>% slice_min(sd, n = 10)
```

```{r}
sid <- 'EP694646'

ps_obs <- momspi %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~max(.) > 1e-3)
ps_cal <- momspi_cal %>%
  filter_sample_data(host_subject_id == sid) %>%
  transform_sample_counts(close_elts) %>%
  prune_taxa(taxa_names(ps_obs), .)
x <- list(Calibrated = ps_cal, Observed = ps_obs) %>%
  map_dfr(as_tibble, .id = "type") %>%
  mutate(across(.otu, fct_relabel, function(x) str_replace_all(x, '_', ' ')))
ntaxa(ps_obs)
```

```{r trajectories-within-subject-low-var, fig.dim = c(6, 9)}
p_sp <- x %>%
  mutate(.otu = fct_reorder(.otu, -.abundance)) %>%
  ggplot(aes(host_visit_number, .abundance, color = type)) +
  scale_y_log10() +
  facet_wrap(~.otu, ncol = 1, scales = 'free_y') +
  geom_line(aes(group = type)) +
  labs(y = 'Proportion', title = 'Species proportions') +
  scale_color_brewer(type = 'qual', palette = 6) +
  theme(legend.position = 'top')
p_me <- mean_eff1 %>%
  filter(host_subject_id == sid) %>%
  ggplot(aes(host_visit_number, mean_efficiency)) +
  scale_y_log10() +
  labs(y = 'Efficiency', title = 'Mean efficiency') +
  geom_line(aes(group = 1))
p_me / p_sp &
  labs(x = 'Host visit') &
  plot_layout(heights = c(0.1, 1)) &
  plot_annotation(tag_levels = 'A')
```

Note that the variation in mean efficiency is very small.


# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>
