---
title: "Explore gut datasets"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - R
date: 2021-11-10
draft: false
output:
  distill::distill_article:
    self_contained: false
    dev: svg
---


```{r}
library(here)
library(tidyverse)
library(speedyseq)
library(metacal)

library(cowplot)
theme_set(theme_cowplot())
```


```{r}
# curatedMetagenomicData("HMP")
tse <- curatedMetagenomicData::curatedMetagenomicData(
  "HMP_2019_ibdmdb.relative_abundance", 
  dryrun = FALSE
)
stopifnot(identical(length(tse), 1L))
tse <- tse[[1]]
ibd_ps <- mia::makePhyloseqFromTreeSummarizedExperiment(
  tse,
  abund_values = "relative_abundance"
) %>%
  rename_with_sample_data(janitor::make_clean_names) %>%
  rename_with_tax_table(janitor::make_clean_names)
sam <- ibd_ps %>% sample_data %>% as_tibble
```

```{r}
```

## metadata

```{r}
sam %>% glimpse
sam %>% dplyr::count(body_site)
sam %>% dplyr::count(dna_extraction_kit)
sam %>% dplyr::count(location)
sam %>% dplyr::count(subject_id) %>% dplyr::count(n)
sam %>% dplyr::count(study_condition, disease, disease_subtype)
```

Most subjects have many samples.



## what is dominance like?

```{r}
ps <- ibd_ps %>% transform_sample_counts(close_elts)
x <- ps %>% as_tibble
```

```{r}
x %>%
  group_by(.sample) %>%
  slice_max(.abundance, n = 1) %>%
  ggplot(aes(.abundance)) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x %>%
  group_by(.sample) %>%
  slice_max(.abundance, n = 1) %>%
  pull(.abundance) %>%
  summary
```

Note that even in the gut, the most abundant species may often be dominant.


## B:F ratio

```{r}
ps.phy <- ps %>% tax_glom('phylum')
x.phy <- ps.phy %>% as_tibble
```

```{r}
x.phy %>%
  group_by(.sample) %>%
  slice_max(.abundance, n = 1) %>%
  pull(.abundance) %>%
  summary
```

```{r}
x.phy %>%
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  ggplot(aes(.abundance + 1e-4, fill = disease_subtype)) +
  facet_wrap(~phylum, ncol = 1) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x.phy %>%
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  mutate(
    across(.abundance, ~ . + 1e-4),
    ratio_f = .abundance / .abundance[phylum == 'Firmicutes']
  ) %>%
  filter(phylum == 'Bacteroidetes') %>%
  ggplot(aes(ratio_f, fill = disease_subtype)) +
  scale_x_log10(breaks = 10^(-3:4)) +
  facet_wrap(~disease, ncol = 1) +
  geom_histogram() +
  theme_minimal_vgrid()
```

There is a lot of variation, but most if it is generally not associated with disease.
This suggests that in this case, the main impact of variation in mean efficinecy from B:F ratio may be to create noise.

(note, I am ignoring structure in the data such as replicate sampling)

how correlated are species within a phylum?
what is the typical within-phylum diversity?

Question: is the HMP2 gut protocol the same as HMP1?


What is the inverse simpson diversity within phyla?

```{r}
div <- x %>% 
  filter(phylum %in% c('Bacteroidetes', 'Firmicutes')) %>%
  with_groups(c(.sample, phylum), mutate,
    across(.abundance, close_elts)
  ) %>%
  with_groups(c(.sample, phylum), summarize,
    diversity = 1/sum(.abundance^2)
  )
```

```{r}
div %>%
  ggplot(aes(diversity)) +
  scale_x_log10() +
  facet_wrap(~phylum, ncol = 1) +
  geom_histogram() +
  theme_minimal_vgrid()
```

```{r}
div %>%
  with_groups(phylum, summarize,
    qs = quantile(diversity, c(0.25, 0.5, 0.75), na.rm = TRUE), 
    prob = c(0.25, 0.5, 0.75)
  )
```

