---
title: "Perform Leopold and Busby (2020) priority-effects analysis with and without bias correction"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:leopold2020host
  - differential abundance
  - bias sensitivity
date: 2022-01-05
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE
  # autodep = TRUE,
  # cache.comments = FALSE
)
```

This Rmd determines whether bias correction has an impact on the analysis of priority effects in @leopold2020host.

# Run analyses with and without bias correction

```{r libraries}
library(here)

set.seed(42)
```

https://github.com/dleopold/Populus_priorityEffects

This script follows the analysis of `code/priorityEffects.R`

Start by running the setup of libraries and loading the phyloseq object and bias estimate.
The code is verbatim from the @leopold2020host code except for the changes in directory necessary for the calls to `source()` and file reading.

```{r lp-setup}
# Analysis of the relative advantage of preemptive colonization.

library(tidyverse)
library(magrittr)
library(phyloseq)
library(foreach)
library(doMC)
library(ggthemes)

this_dir <- getwd()
setwd(here('notebook/_data/leopold2020host/dleopold-Populus_priorityEffects-8594f7c/'))

source("code/Rfunctions.R")
source("code/colors.R")

# register cores for parallel processing
registerDoMC(parallel::detectCores())

# load phyloseq data
(phy <- loadPhyloseq())

# identify focal taxa for plotting 
focalTaxa <- unique(sample_data(phy)$Treatment)

# Extract OTU data, convert to proportions and long format
bias <- read.csv("output/tabs/bias.csv")

setwd(this_dir)
```

Next, we wrap the analyses of the original study within a function `analyze_priority_effects()` to facilitate running the same analyses on the calibrated and uncalibrated profiles.
The code in the function is copied verbatim from the source in `code/priorityEffects.R` except that the call to `unbias()` in the very first line is skipped so that we can control whether correction is done.
Click the toggle to expand the code chunk.

```{r, code_folding = TRUE}
analyze_priority_effects <- function(phy, nboots = 10000) { 
  df <- phy %>%
    transform_sample_counts(function(x){x/sum(x)}) %>%
    otu_table %>% data.frame %>%
    dplyr::select(all_of(focalTaxa)) %>%
    bind_cols(sample_data(phy) %>% data.frame %>% 
                dplyr::select(Region,Genotype,Treatment)) %>%
    pivot_longer(all_of(focalTaxa),names_to="Taxa",values_to="proportion") %>%
    mutate(Focal=ifelse(Taxa==Treatment,T,F),
           Region=ifelse(Region=="East","E","W") %>%
             factor(levels=c("W","E")))

  # Define a function to calculate the estimated priority effect strength for each species on each host genotype as the log-ratio of the proportional abundance when arriving early vs not arriving early
  getPEs <- function(df) {
    df %>% group_by(Focal,Region,Taxa,Genotype,Treatment) %>%
      summarize_all(gm_mean) %>% ungroup %>%
      group_by(Region,Genotype,Taxa) %>%
      summarise(PE=log(mean(proportion[Focal])/mean(proportion[!Focal])))
  }
  meanPEs <- getPEs(df) 
  meanPEs$Genotype %<>% factor(.,levels=unique(.))

  #####################
  ### Bootstrap CIs ###
  #####################

  # Get bootstrap confidence intervals on point estimates of priority effects for each species on each genotype
  # nboots <- 10000
  bootPEs <- foreach(i=1:nboots, .combine=bind_rows) %dopar% {
    df %>% group_by(Region, Taxa, Genotype,Focal,Treatment) %>%
      sample_frac(replace=T) %>%
      getPEs() %>% mutate(bootID=i)
  }
  bootPEs$Genotype %<>% factor(.,levels=unique(bootPEs$Genotype))

  # Get bias corrected and accelerated confidence intervals
  bootPE.ci <- bootPEs %>% 
    group_by(Genotype,Taxa) %>%
    summarize(LCI=coxed::bca(PE)[1],
              UCI=coxed::bca(PE)[2]) %>% full_join(meanPEs)

  ###################
  ### Region test ###
  ###################

  # First get the mean priority effect for each species on eastern and western genotypes
  regionPEs <- meanPEs %>% group_by(Region,Taxa) %>%
    summarize(meanPE=mean(PE),
              tstat=t.test(PE,mu=0,alternative="greater") %>% .$statistic)

  # Run bootstrapped t-test using 100 bootstraps at the region level for each of the previously generated genotype-level bootstraps
  bootRegion <- foreach(i=1:100,.combine=bind_rows) %dopar% {
    bootPEs %>% 
      group_by(Region,Taxa,bootID) %>%
      sample_frac(replace=T) %>%
      left_join(regionPEs,by=c("Region","Taxa")) %>%
      mutate(center=PE-meanPE) %>%
      summarize(mu.boot=mean(PE),
                tstat.obs=mean(tstat),
                tstat.boot=ifelse(var(center)==0,ifelse(mean(PE)>mean(meanPE),Inf,-Inf),
                                  t.test(center,mu=0) %>% .$statistic))}
  # calculate bootstrapped p-values (one-tailed test for significant positive priority effects)
  bootRegionPvals <- bootRegion %>% drop_na %>%
    mutate(test = tstat.boot+1>tstat.obs+1) %>% 
    summarize(pval=mean(as.numeric(test))) %>%
    mutate(stars=gtools:::stars.pval(pval) %>% gsub(".","+",fixed = T,.)) 
  regionSig <- bootRegion %>% summarize(mu.boot.max=max(mu.boot)) %>%
    left_join(bootRegionPvals)

  ############
  ### PLOT ###
  ############

  #' Define a function to plot the regional priority effects bootstrap results 
  get_inset <- function(df,inset.ymin,inset.ymax){
    ggplot(df, 
           aes(x=Region,y=mu.boot)) +
      geom_violin(scale="width",fill="grey75",color="grey75") + 
      geom_text(data=df %>% dplyr::select(Region,Taxa,stars,mu.boot.max) %>%
                  slice(1),
                aes(label=stars,y=mu.boot.max),size=5)+
      geom_hline(yintercept = 0,alpha=0.4,linetype="dotted")+
      coord_cartesian(clip='off')+
      ylim(c(inset.ymin,inset.ymax))+
      theme_few()+
      theme(strip.text = element_blank(),
            axis.title = element_blank(),
            panel.border = element_blank(),
            axis.line = element_line(color="black",size=0.2))
  }

  # create inset plots as a list
  insets <- bootRegion %>% 
    left_join(dplyr::select(regionSig,Region,Taxa,stars,mu.boot.max)) %>%
    split(f = .$Taxa) %>%
    purrr::map(~annotation_custom2(
      grob = ggplotGrob(get_inset(.,min(bootRegion$mu.boot),max(bootRegion$mu.boot))), 
      data = data.frame(Taxa=unique(.$Taxa)),
      ymin = 2.4, ymax=Inf, xmin=4, xmax=Inf)
    )

  # make full plot of bootstrapped priority effects results
  plt <- ggplot(bootPE.ci,aes(x=Genotype,y=PE))+
    geom_pointrange(aes(ymin=LCI,ymax=UCI,fill=Genotype,shape=Genotype),
                    size=0.5,stroke=0.35,fatten=5.5)+
    scale_shape_manual("Host genotype", values=c(rep(21,7),rep(23,5)))+
    scale_fill_manual("Host genotype",values=pal.genotype)+
    geom_hline(yintercept = 0,alpha=0.6,linetype="dotted")+
    scale_x_discrete(breaks = levels(bootPEs$Genotype),
                     limits = c(levels(bootPEs$Genotype)[1:7], "skip1",
                                levels(bootPEs$Genotype)[8:12]),
                     expand=expansion(0.075))+
    ylab("Strength of priority effect")+
    labs(tag="Eastern              Western")+
    coord_cartesian(clip='off')+
    ylim(c(-1,3.5))+
    facet_wrap(~Taxa,scales="free_x",nrow=1)+
    guides(fill = guide_legend(override.aes = list(size=0.8)))+
    theme_few()+
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          strip.text = element_text(size=14,face="italic"),
          axis.title.y = element_text(size=14),
          legend.title = element_text(size=14),
          legend.text = element_text(size=12),
          legend.margin=margin(6,6,6,12),
          plot.tag = element_text(angle=90,hjust=0),
          plot.tag.position = c(0.848,0.115))+
    insets
  # ggsave("output/figs/Fig.3.pdf",width=24,height=10,units="cm")

  return(
    list(
      meanPEs = meanPEs, 
      bootPEs = bootPEs,
      bootPE.ci = bootPE.ci,
      regionPEs = regionPEs,
      bootRegion = bootRegion,
      bootRegionPvals = bootRegionPvals,
      plot = plt
    )
  )
}
```

Next, we run the analysis on the phyloseq objects with and without bias correction (calibration).
@leopold2020host use a custom function `unbias()` for bias correction, which also implements a zero-replacement procedure.
In order to keep the calibrated and uncalibrated analyses comparable, we should apply the same zero-replacement procedure in both cases.
We do this by applying `unbias()` with the supplied bias vector set to all 1s, which corresponds to no bias.

```{r}
bias_list <- list(
  'Corrected' = bias, 
  'Uncorrected' = bias %>% mutate(Bhat = 1)
)
phy_list <- bias_list %>% map(~unbias(phy, .x))
```

The only difference between the two phyloseq objects should be a perturbation by the bias vector, which we can check by using `metacal::calibrate()`.

```{r}
tmp <- metacal::perturb(
  phy_list$Corrected, 
  bias %>% filter(Taxon != 'Melampsora') %>% select(Taxon, Bhat) %>% deframe,
  norm = 'keep'
)
list(phy_list$Uncorrected, tmp) %>%
  map(otu_table) %>%
  map(transform_sample_counts, function(x) x / sum(x)) %>%
  {all.equal(.[[1]], .[[2]])}
```

The sample totals are affected by `unbias()` such that they differ between the corrected and uncorrected objects,

```{r}
phy_list$Corrected %>% sample_sums %>% head(3)
tmp %>% sample_sums %>% head(3)
phy_list$Uncorrected %>% sample_sums %>% head(3)
```

But since the analysis function converts the counts to proportions in the first step, this difference can't impact the results.

Finally, we run the analysis on both the corrected and uncorrected phyloseq objects.
The call is wrapped in `xfun::cache_rds()` to cache the results; the hash is based on the inputs and the analysis function's source code.

```{r analyses, message = FALSE, results = 'hide'}
fun_src <- attr(analyze_priority_effects, 'srcref') %>% as('character')
res <- xfun::cache_rds({
  phy_list %>% map(analyze_priority_effects)
}, hash = list(fun_src, phy_list, focalTaxa),
  clean = FALSE
)
```

# Compare results

Next we compare the results of the analysis with and without bias correction.

```{r}
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())
```

```{r}
res %>% names
res[[1]] %>% names
```

First, let's compare the main text figure with and without correction,

```{r, fig.dim = c(10, 8), layout="l-body-outset"}
map2(map(res, 'plot'), names(res), ~.x + ggtitle(.y)) %>%
  {.[[1]]  / .[[2]]} &
  theme(plot.title.position = 'plot')
```

The results are nearly indistinguishable; bias correction has no significant impact.

Next, let's compare the point estimates of the priority effects,

```{r, fig.dim = c(6, 6)}
x <- res %>%
  map_dfr('meanPEs', .id = 'type') %>%
  pivot_wider(names_from = type, values_from = PE)
mse <- x %>%
  ungroup %>%
  summarize(
    mse = mean((Corrected - Uncorrected)^2)
  ) %>% .[[1]] %>% signif(3)
x %>%
  ggplot(aes(Uncorrected, Corrected, color = Region)) +
  coord_fixed() +
  geom_abline(color = 'darkred') +
  geom_point() +
  labs(title = 'Effect of bias correction on priority-effect estimate', 
    subtitle = str_glue('Mean squared difference: {mse}')
  ) +
  theme(plot.title.position = 'plot') +
  scale_color_brewer(type = 'qual', palette = 2)
```

There is very little impact on the point estimates.
Now, let's compare the bootstrap confidence intervals with and without correction,

```{r, fig.dim = c(8, 8)}
res %>%
  map_dfr('bootPE.ci', .id = 'type') %>%
  ggplot(aes(x = Genotype, y = PE, color = type, ymin = LCI, ymax =UCI)) +
  facet_wrap(~Taxa, scales = "fixed", ncol = 1) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  scale_color_brewer(type = 'qual') +
  theme(
    plot.title.position = 'plot',
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(title = 'Effect of bias correction on priority-effect CIs')
```

This fig clearly shows the equivalence of results with uncorrected and corrected taxonomic profiles.

Let's do similar for the regional PEs.

```{r, fig.dim = c(7, 4.5)}
x <- res %>%
  map_dfr('regionPEs', .id = 'type') %>%
  pivot_longer(c(meanPE, tstat)) %>%
  pivot_wider(names_from = type)
mse <- x %>%
  ungroup %>%
  filter(name == 'meanPE') %>%
  summarize(
    mse = mean((Corrected - Uncorrected)^2)
  ) %>% .[[1]] %>% signif(3)
x %>%
  ggplot(aes(Uncorrected, Corrected)) +
  facet_wrap(~name, scales = 'free') +
  # coord_fixed() + # doesn't work with free scales
  geom_abline(color = 'darkred') +
  geom_point() +
  labs(title = 'Effect of bias correction on regional priority-effect estimate', 
    subtitle = str_glue('Mean squared difference in meanPE: {mse}')
  ) +
  theme(plot.title.position = 'plot')
```

Again there is almost no effect, except that the largest t statistic is somewhat larger before bias correction.

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>
