---
title: >
  Exploring the effect of bias in Leopold and Busby (2020)
author: Michael McLaren
date: '2019-10-21'
categories:
  - bias sensitivity
  - R
  - ref:leopold2020host
description:
bibliography: _references.bib
draft: false
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    dev: svg
---

```{r, include = FALSE}
# knitr chunk options
knitr::opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

Analysis of bias in @leopold2020host.

# Setup

```{r setup}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
theme_set(theme_cowplot())

# stats helpers
library(broom)

library(metacal); packageVersion("metacal")
```

```{r load-data, include = FALSE}
## From the metacal 2.0 tutorial
# Download data from https://zenodo.org/record/3872145
data_path <- here::here("_data", "leopold2020host")
# To use a temporary directory:
# data_path <- file.path(tempdir(), "leopold2020")
if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  download.file(
    "https://zenodo.org/record/3872145/files/dleopold/Populus_priorityEffects-v1.2.zip",
    file.path(data_path, "Populus_priorityEffects-v1.2.zip")
  )
  unzip(
    file.path(data_path, "Populus_priorityEffects-v1.2.zip"), 
    exdir = data_path
  )
}
#> The microbiome data is stored in a phyloseq object,
ps <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/output/compiled/phy.rds") %>%
  readRDS

sample_data(ps) <- sample_data(ps) %>%
  transform(
    Timepoint = factor(Timepoint)
  )

mock_actual <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/data/MockCommunities.csv") %>%
  read.csv(row.names = 1) %>%
  select(-Sym4) %>%
  as("matrix") %>%
  otu_table(taxa_are_rows = FALSE) %>%
  transform_sample_counts(function(x) close_elts(1 / x))
mock_taxa <- taxa_names(mock_actual)
sam <- sample_data(ps) %>% as("data.frame") %>% as_tibble(rownames = "Sample")
tax <- tax_table(ps) %>% as("matrix") %>% as_tibble(rownames = "Taxon")
ps.mock <- ps %>% 
  subset_samples(Samp_type == "Mock") %>%
  prune_taxa(mock_taxa, .)
```

```{r estimate-bias}
ps.mock.pseudo <- ps.mock %>%
  transform_sample_counts(function(x) x + 1)
mc_fit <- estimate_bias(ps.mock.pseudo, mock_actual, boot = TRUE) %>% print
bias <- coef(mc_fit) %>% print
```

```{r}
mc_fit.summary <- summary(mc_fit)
print(mc_fit.summary)
```

```{r, fig.dim = c(5,3)}
coef_tb <- mc_fit.summary$coefficients
coef_tb %>%
  mutate(taxon = fct_reorder(taxon, estimate)) %>%
  ggplot(aes(taxon, estimate, 
      ymin = estimate / gm_se^2, ymax = estimate * gm_se^2)) +
  geom_hline(yintercept = 1, color = "grey") +
  geom_pointrange() +
  scale_y_log10() +
  coord_flip()
```

```{r calibrate}
ps.pseudo <- transform_sample_counts(ps, function(x) x + 1)
ps.pseudo.cal <- calibrate(ps.pseudo, bias) %>% print
```


# Sample mean efficiency across samples

To estimate the SMRE with formula `sum_i A_i E_i`, use the calibrated abundances.

```{r}
ps0 <- ps.pseudo.cal %>%
  prune_taxa(names(bias), .)
```

```{r}
otu <- ps0 %>% otu_table %>% orient_taxa(as = "cols") %>%
  transform_sample_counts(close_elts)
sam0 <- sample_data(ps0) %>% 
  as("data.frame") %>% 
  as_tibble(rownames = "Sample")
```

```{r}
sme <- otu %>% psmelt %>%
  left_join(bias %>% enframe("OTU", "Efficiency"), by = "OTU") %>%
  group_by(Sample) %>%
  summarize(SMRE = sum(Efficiency * Abundance)) %>%
  left_join(sam0, by = "Sample") %>%
  mutate(
    across(Timepoint, as.factor),
    across(Timepoint, fct_explicit_na, "Other"),
  )
```

```{r}
range(sme$SMRE)
range(bias)
```


```{r}
p1 <- ggplot(sme, aes(SMRE, fill = Samp_type)) +
  geom_histogram() +
  annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  expand_limits(x = range(bias)) +
  scale_fill_brewer(type = "qual") +
  labs(
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency", 
    y = "Number of samples",
    fill = "Sample type"
  )
p1 +
  plot_annotation(
    title = "Distribution of sample mean relative efficiencies",
  )
```

```{r, eval = F}
ggsave("/tmp/leopold2020host-smre-distribution.svg", units = "in",
  width = 6, height = 3, scale = 1.5)
ggsave("/tmp/leopold2020host-smre-distribution.png", units = "in",
  width = 6, height = 3, scale = 1.5)
```

In this case, the bimodal distribution is likely driven by the high proportion
of Melampsora (pathogen) in ~half the samples.

TODO: check this hypothesis. by making a histogram of melampspora %, and a scatterplot of the melamspora vs. SMRE. Can do with facet matrix from ggforce

```{r}
mel <- otu %>%
  prune_taxa("Melampsora", .) %>%
  psmelt %>%
  select(-OTU, Melampsora = Abundance)
sme0 <- left_join(sme, mel, by = "Sample")
```

```{r}
# p2 <- ggplot(sme0, aes(SMRE, fill = Melampsora)) +
p2 <- ggplot(sme0, aes(y = Melampsora, x = SMRE, color = Samp_type)) +
  geom_point() +
  # annotate("point", y = 1, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  # scale_y_sqrt() +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 1e-2),
    breaks = c(0, 0.01, 0.1, 0.5, 1)
    ) +
  expand_limits(x = range(bias)) +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Melampsora proportion",
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency",
    color = "Sample type"
  )
p2
```

```{r}
(p1 + ggtitle("Distribution of the mean efficiency across samples")) / 
  (p2 + ggtitle("Melampsora proportion vs. mean efficiency")) +
  plot_annotation(tag_levels = "A")
```

```{r, eval = F}
ggsave("/tmp/leopold2020host-smre-distribution.svg", units = "in",
  width = 6, height = 6, scale = 1.5)
ggsave("/tmp/leopold2020host-smre-distribution.pdf", units = "in",
  width = 6, height = 6, scale = 1.5)
ggsave("/tmp/leopold2020host-smre-distribution.png", units = "in",
  width = 6, height = 6, scale = 1.5)
```

Note, the high-Mel samples are the timepoint 2 samples, and were not used for some (all?) of his analysis due to the dominance of Mel. Let's check this out,

```{r}
sme %>%
  mutate(
    across(Timepoint, as.factor),
    across(Timepoint, fct_explicit_na, "Other"),
  ) %>%
  # filter(Samp_type == "Experiment") %>%
  ggplot(aes(SMRE, fill = Timepoint)) +
  geom_histogram() +
  annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  expand_limits(x = range(bias)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  )
```


```{r}
sme %>%
  filter(Samp_type == "Experiment") %>%
  ggplot(aes(SMRE, fill = Treatment)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

```{r}
sme %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Region)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

Can see that Eastern genotypes more susceptible (more Mel and higher SME)


Are some western genotypes less susceptible?

```{r}
sme %>%
  filter(Samp_type == "Experiment", Region == "West",
    Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Genotype)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

Eastern?

```{r}
sme %>%
  filter(Samp_type == "Experiment", Region == "East",
    Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Genotype)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    x = "Mean efficiency", 
    y = "Number of samples"
  ) +
  facet_wrap(~Timepoint, ncol = 1)
```

### Alt plot

Might be better to just focus on the treated experimental samples; can note
that the bimodal distribution in SMRE is due to the presence of Mel in the
timepoint 2, and thus we'd mess up our AA inferences if we tried to infer the
change in AA after Mel introduction. We might also (if were were analyzing
timepoint 2 samples) 

Actually, including the bias values serves the important purpose of showing how
the SME varies less than the efficiencies. For that reason it could even be
worth having the single-isolate samples, though seems unneeded.

Points

* SME varies less across samples than efficiencies vary across taxa
* SME much greater after rusting (timepoint 2) -> it will look like the growth of the
  pathogen suppressed the proportion and absolute abundance of the colonizers
  5-6X more than it actually did
* The Mel prop and hence the SME varies significantly in the timepoint 2
  samples. This may lead to systematic errors in certain inferences. E.g., if
  host genotype affects Mel and another taxon, that other taxon's value will be
  distorted (and, I think, the effect on Mel will be inflated).
  * As an example, think through the case of more resistant host genotypes
  * Might also consider the taxa found to suppress Mel; see paper

```{r}
p1.1 <- sme %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  ggplot(aes(SMRE, fill = Timepoint)) +
  geom_histogram() +
  annotate("point", y = 200, x = bias, color = "black", shape = "|", size = 7) +
  expand_limits(x = range(bias)) +
  scale_x_log10() +
  scale_y_sqrt(breaks = c(1, 10, 50, 100, 200)) +
  scale_fill_brewer(type = "qual") +
  labs(
    # x = "Sample mean relative efficiency (SMRE)",
    x = "Mean efficiency", 
    y = "Number of samples"
  )
p2.1 <- sme0 %>%
  filter(Samp_type == "Experiment", Treatment != "Negative") %>%
  ggplot(aes(y = Melampsora, x = SMRE, color = Timepoint)) +
  geom_point() +
  expand_limits(x = range(bias)) +
  scale_x_log10() +
  # scale_y_sqrt() +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(sigma = 1e-2),
    breaks = c(0, 0.01, 0.1, 0.5, 1)
    ) +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Melampsora proportion",
    x = "Mean efficiency"
  )
```

```{r}
(p1.1 + ggtitle("Distribution of the mean efficiency across samples")) / 
  (p2.1 + ggtitle("Melampsora proportion vs. mean efficiency")) +
  plot_annotation(tag_levels = "A")
```

TODO: Still need to determien if the y axis is accurate on histogram; I think
its off when using pseudo-log. But perhaps can get away with linear?


## Check errors in FCs

We can use this distribution to pick examples of samples where bias
approximately cancels in the FC estimates (because their SMRE's are similar),
and samples where bias does not approximately cancel (because their SMRE's are
very different). Could do for mock samples (using the ground truth) or for
natural samples (using the calibrated composition). Might as well do it both
ways; not sure which would be best for a paper. The mocks have the benefit of
knowing the ground truth, while the natural samples demonstrate practical
significance. Also, we can give nice intuitive description of what we're doing
in the natural samples by saying that we're comparing samples w/ low-low,
high-high, or low-high pathogen %.


TODO ^^

### mock samples


To pick from the mock samples:

```{r}
sme0 %>%
  filter(Samp_type == "Mock") %>%
  select(Sample, SMRE, Melampsora) %>%
  arrange(SMRE) %>%
  mutate(
    log10_SMRE = log10(SMRE),
    diff = c(NA, diff(log10_SMRE))
  )
```

- Mock.3 -> Mock.5 have the biggest difference in SMRE. Notably, its still only
  a factor of 5.5 difference.
- Mock.7 -> Mock.9 have the smallest log difference.

Next: Compute the actual and observed FC in proportions, for each of these
pairs of communities. Perhaps define a function to spit out the relevant data
frame, given any pair of samples, the phyloseq object (or calibrated phyloseq
object?), and the bias vector? Actually, for the mocks, should use the
"actual" table instead of calibrated. 

Use these objects,

```{r}
mock_actual
ps.mock.pseudo
```

We can use `metacal::pairwise_ratios()` to compute the FCs in abundances
between all samples + for all taxa, given a phyloseq object.

```{r}
fcs <- list(Actual = mock_actual, Observed = ps.mock.pseudo) %>%
  map(transform_sample_counts, close_elts) %>%
  map(pairwise_ratios, margin = "samples", filter = FALSE) %>%
  map_dfr(psmelt, .id = "Type") %>%
  # redo sample.1 and sample.2 since missing for the mock
  select(-sample.1, -sample.2) %>%
  rename(Pair = Sample, Fold_change = Abundance) %>%
  separate(Pair, str_c("Sample.", 1:2), sep = ":", remove = FALSE) %>%
  mutate(across(OTU, factor, levels = bias %>% sort %>% names)) %>%
  left_join(sme0 %>% select(Sample.1 = Sample, SMRE, Melampsora), 
    by = "Sample.1") %>%
  left_join(sme0 %>% select(Sample.2 = Sample, SMRE, Melampsora), 
    by = "Sample.2", suffix = c(".1", ".2"))
```

```{r}
fcs %>%
  filter(Pair %in% c("Mock.5:Mock.3", "Mock.9:Mock.7")) %>%
  ggplot(aes(y = OTU, x = Fold_change, shape = Type, color = Type)) +
  facet_grid(. ~ Pair) +
  scale_shape_manual(values = c(Actual = 3, Observed = 1)) +
  scale_color_brewer(type = "qual", palette = 6) +
  geom_vline(xintercept = 1, color = "black", size = 0.2) +
  geom_point(size = 2) +
  theme_minimal_hgrid() +
  scale_x_log10(limits = c(3e-3, 1e2), oob = scales::squish) +
  labs(x = "Fold change in proportion") +
  plot_annotation(
    title = "Actual and observed fold changes between pairs of mocks"
  )
```

```{r}
ggsave("/tmp/leopold2020-fc-mock-examples.png", units = "in",
  width = 6, height = 3, scale = 1.5)
```


Can see that on the left, the error is positive for all taxa, and of a similar
magnitude (on the order of ~6X, as predicted) for all taxa except for
Epicoccum, which dropped out during sequencing in Mock.5; and on the right, the
error is smaller and not consistently positive or negative.  Sign errors
correspond to the observed and actual fold changes being on opposing sides of
the vertical red line at 1.

todo: add the SMRE changes to the facet labels, and perhaps mark the expected 

What other types of plots, which might use all pairs of samples?

- we could plot the observed vs. predicted error

- we could check that the error really is taxon independent; one way to do this
  is to calc the correlation in LFC between observed and actual

```{r}
lvls <- sme0 %>% arrange(SMRE) %>% pull(Sample)
fcs0 <- fcs %>%
  # filter(Sample.1 < Sample.2) %>%
  select(Type:Fold_change, SMRE.1:Melampsora.2) %>%
  pivot_wider(names_from = Type, values_from = Fold_change) %>%
  mutate(across(c(Sample.1, Sample.2), factor, levels = lvls))
```

```{r}
fcs0 %>%
  filter(as.integer(Sample.1) < as.integer(Sample.2)) %>%
  ggplot(aes(Actual, Observed, color = OTU)) +
  geom_abline(color = "grey") +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  facet_grid(Sample.1 ~ Sample.2) +
  scale_color_brewer(type = "qual", palette = 3)
```

Note how (especially) for Mock 1 and 6, the points follow a straight line with
a greater slope than predicted. This seems likely related to the systematic
deviation from the model that I observed. 

Check that there is indeed very high correlations; remove the Epicoccum Mock-5
dropout first to avoid major distortions just from that.

```{r}
corrs <- fcs0 %>%
  filter(
    Sample.1 != Sample.2,
    !((Sample.1 == "Mock.5" | Sample.2 == "Mock.5") & OTU == "Epicoccum")
  ) %>%
  mutate(across(c(Actual, Observed), log10)) %>%
  group_by(Pair) %>%
  summarize(correlation = cor(Actual, Observed, method = "pearson"))
corrs %>% pull(correlation) %>% summary
# qplot(corrs$correlation)
```

### Experiment samples

We could just try picking a high and low Melampsora sample - perhaps from the
mode of the peaks, rather than the extreme.

```{r}
sme0 %>%
  filter(Samp_type == "Experiment") %>%
  select(Sample, SMRE, Melampsora) %>%
  arrange(SMRE) %>%
  mutate(
    log10_SMRE = log10(SMRE),
    diff = c(NA, diff(log10_SMRE))
  )
```

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>
